---
title: "Group X: NYC Green Cab"
author: 'Jennifer Lieu, Gehua Zhang, Xinquan Wang, Yanzhi Zhang'
date: "April 25 2019"
output: slidy_presentation
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

```{r library, include=FALSE, warning=FALSE}
library(data.table)
library(ggplot2)
library(dplyr)
library(purrr)
library(tidyr)
library(lubridate)
library(corrplot)
library(gridExtra)
```

```{r read data+constant, echo=FALSE, warning=FALSE}
green <- fread(input = '../Data/2018_Green_Taxi_Trip_Data.csv',verbose = FALSE,na.strings=c(""))
weather <- fread(input = '../Data/weather_new.csv',verbose = FALSE,na.strings=c(""))

weather <- na.omit(weather)
```

# Why This Topic?

- Assisting Cab Drivers

- Popular Locations helps both parties

- Determining Revenues helps smart business decisions

# Data Sets

## NYC Green Cab Dataset

- January 2018 to December 2018

- 19 Variables

- 8807303 observations

- Link to data set https://data.cityofnewyork.us/Transportation/2018-Green-Taxi-Trip-Data/w7fs-fd9i

## Weather data

- January 2018 to December 2018

- 124 Variables

- 8760 observations (missing a few hours' records)

- Link to data set https://www.ncdc.noaa.gov/cdo-web/search

# Introduction to the Problem: 

We used the 2018 NYC green cab data provided by NYC open data which is a government-run organization.

```{r}
head(green)
```

- **VendorID:** A code indicating the LPEP provider that provided the record.
1= Creative Mobile Technologies, LLC; 2= VeriFone Inc.
- **lpep_pickup_datetime:** The date and time when the meter was engaged.
- **lpep_dropoff_datetime:** The date and time when the meter was disengaged.
- **Passenger_count:** The number of passengers in the vehicle. This is a driver-entered value.
- **Trip_distance:** The elapsed trip distance in miles reported by the taximeter.
- **PULocationID:** TLC Taxi Zone in which the taximeter was engaged
- **DOLocationID:** TLC Taxi Zone in which the taximeter was disengaged
- **RateCodeID**The final rate code in effect at the end of the trip.
1= Standard rate
2=JFK
3=Newark
4=Nassau or Westchester 5=Negotiated fare 6=Group ride
- **Store_and_fwd_flag** This flag indicates whether the trip record was held in vehicle memory before sending to the vendor, aka “store and forward,” because the vehicle did not have a connection to the server.
Y= store and forward trip
N= not a store and forward trip
- **Payment_type:** A numeric code signifying how the passenger paid for the trip.
1= Credit card 2= Cash
3= No charge 4= Dispute
5= Unknown 6= Voided trip
- **Fare_amount:** The time-and-distance fare calculated by the meter.
- **MTA_tax:** $0.50 MTA tax that is automatically triggered based on the metered rate in use.
- **Improvement_surcharge:** $0.30 improvement surcharge assessed on hailed trips at the flag drop. The improvement surcharge began being levied in 2015.
- **Tip_amount:** Tip amount – This field is automatically populated for credit card tips. Cash tips are not included.
- **Tolls_amount:** Total amount of all tolls paid in trip.
- **Total_amount:** The total amount charged to passengers. Does not include cash tips.
- **Trip_type:** A code indicating whether the trip was a street-hail or a dispatch that is automatically assigned based on the metered rate in use but can be altered by the driver.
1= Street-hail 2= Dispatch



# Data Cleaning and Visualization

From a quick look in the data description PDFs, we found that "VendorID", "Store_and_fwd_flag", "Payment_type", and "ehail_fee" are not correlated with our problem, so we dropped those columns from the very beginning.

```{r drop_feature}
names(green)
green[, c("VendorID", "store_and_fwd_flag", "payment_type", "ehail_fee")] <- list(NULL)
```

The data was very large (793MB) and messy, we had to clean the data before we performed any data analysis.

# Location ID, Passenger Count

First, we looked at whether the location ID was in the valid range: is 1 to 265.

Quintilian for pick-up location ID
```{r}
quantile(green$PULocationID, probs = c(0, 0.005, 0.025, 0.5, 0.75, 0.975, 0.995, 1))
```
Quintilian for drop-off location ID
```{r}
quantile(green$DOLocationID, probs = c(0, 0.005, 0.025, 0.5, 0.75, 0.975, 0.995, 1))
pruned_green <- green[PULocationID>=0&PULocationID<=266&DOLocationID>=0&DOLocationID<=266,]
```

Usually, we would only have 1 to 4 passenagers in a Sedan. However, there might be more passenger seats in different car models. Since this is a number entered by driver, there could be mistakes in those data entries. Regardless of the possible passengers, we take a 95% confidence interval on the passenger counts.

Quantile for the passenger counts
```{r}
quantile(pruned_green$passenger_count, probs = c(0, 0.005, 0.025, 0.5, 0.75, 0.975, 0.995, 1))
pruned_green <- pruned_green[passenger_count > 0 & passenger_count < 6,]
```
```{r}
pruned_green %>% 
  group_by(passenger_count) %>% 
  count() %>% 
  ggplot(aes(x=passenger_count,y=n, fill=passenger_count))+
  geom_col()+
  theme(legend.position = "none")
```

# Trip Distance 

Quantile for the trip distance
```{r}
quantile(pruned_green$trip_distance, probs = c(0, 0.005, 0.025, 0.5, 0.75, 0.975, 0.995, 1))
pruned_green <- pruned_green[trip_distance > 0 & trip_distance <= 21.17,]
```

```{r}
pruned_green %>% 
  ggplot(aes(x=trip_distance)) + 
  geom_histogram(bins=100, fill="red")+
  theme_bw()+theme(axis.title = element_text(size=11),axis.text = element_text(size=8))+
  ylab("Density")+coord_cartesian(x=c(0,25))
```

# All Kinds of Fares

Quantile for the fare amount
```{r}
quantile(pruned_green$fare_amount, probs = c(0, 0.005, 0.025, 0.5, 0.75, 0.975, 0.995, 1))
pruned_green <- pruned_green[fare_amount >= 2.5 & fare_amount <= 58.0,]
```
Quantile for the extra fees
```{r}
quantile(pruned_green$extra, probs = c(0, 0.005, 0.025, 0.5, 0.75, 0.975, 0.995, 1))
pruned_green <- pruned_green[extra >= 0 & extra <= 1,]
```
Quantile for the mta tax
```{r}
quantile(pruned_green$mta_tax, probs = c(0, 0.005, 0.025, 0.5, 0.75, 0.975, 0.995, 1))
pruned_green <- pruned_green[mta_tax > 0 & mta_tax <= 0.5,]
```
Quantile for the tip amount
```{r}
quantile(pruned_green$tip_amount, probs = c(0, 0.005, 0.025, 0.5, 0.75, 0.975, 0.995, 1))
pruned_green <- pruned_green[tip_amount >= 0 & tip_amount <= 0.5,]
```
Quantile for the tolls amount
```{r}
quantile(pruned_green$tolls_amount, probs = c(0, 0.005, 0.025, 0.5, 0.75, 0.975, 0.995, 1))
pruned_green <- pruned_green[tolls_amount >= 0 & tolls_amount <= 5.76,]
```
Quantile for the improvement surcharge
```{r}
quantile(pruned_green$improvement_surcharge, probs = c(0, 0.005, 0.025, 0.5, 0.75, 0.975, 0.995, 1))
pruned_green <- pruned_green[improvement_surcharge == 0.3, ]
```
Quantile for the total amount
```{r}
quantile(pruned_green$total_amount, probs = c(0, 0.005, 0.025, 0.5, 0.75, 0.975, 0.995, 1))
pruned_green <- pruned_green[total_amount >= 3.3 & total_amount <= 73.32,]
```

# Visualiztion on Fare

We only visualize the fare amount and total amount since the rest are not very influential to the model.
```{r}
plot_1 <- pruned_green %>% 
  ggplot(aes(x=fare_amount)) + 
  geom_histogram(bins=50, fill="red")+
  theme_bw()+theme(axis.title = element_text(size=11),axis.text = element_text(size=8))+
  ylab("Density")+coord_cartesian(x=c(0,60))
plot_3 <- pruned_green %>% 
  ggplot(aes(x=total_amount)) + 
  geom_histogram(bins=50, fill="red")+
  theme_bw()+theme(axis.title = element_text(size=11),axis.text = element_text(size=8))+
  ylab("Density")+coord_cartesian(x=c(0,60))

grid.arrange(plot_1, plot_3, ncol= 2)
```


# Time Feature

The date columns are of type character, so they had to be converted to type numeric. We convert the dates into the seconds elapsed from the first moment of 2018 and group them into time bins by hours. 

```{r data_cleaning_lpep_pickup_datetime_lpep_dropoff_datetime}
one.year <- (as.numeric(strptime("01/01/2019 12:00:00 AM", "%m/%d/%Y %l:%M:%S %p")) - as.numeric(strptime("01/01/2018 12:00:00 AM", "%m/%d/%Y %l:%M:%S %p"))) %/% 3600

pruned_green$pickup_hour <- 
  (as.numeric(strptime(pruned_green$lpep_pickup_datetime, "%m/%d/%Y %l:%M:%S %p")) - as.numeric(strptime("01/01/2018 12:00:00 AM", "%m/%d/%Y %l:%M:%S %p"))) %/% 3600

pruned_green$dropoff_hour <- (as.numeric(strptime(pruned_green$lpep_dropoff_datetime, "%m/%d/%Y %l:%M:%S %p")) - as.numeric(strptime("01/01/2018 12:00:00 AM", "%m/%d/%Y %l:%M:%S %p"))) %/% 3600
```

Quantile for the pick-up hours
```{r}
quantile(pruned_green$pickup_hour, probs = c(0, 0.005, 0.025, 0.5, 0.75, 0.975, 0.995, 1))
```
Quantile for the drop-off hours
```{r}
quantile(pruned_green$dropoff_hour, probs = c(0, 0.005, 0.025, 0.5, 0.75, 0.975, 0.995, 1))

pruned_green <- pruned_green[pickup_hour>=0 & pickup_hour<=one.year,]
pruned_green <- pruned_green[dropoff_hour>=0 & dropoff_hour<=one.year,]
```

# Feature Engineering

We think trip duration can be an expressive feature for this model, so we calculated the duration simply by subtracting drop-off hour by pickup-hour. 

```{r}
pruned_green$trip_duration <- 
  ((as.numeric(strptime(pruned_green$lpep_dropoff_datetime, "%m/%d/%Y %l:%M:%S %p")) - as.numeric(strptime("01/01/2018 12:00:00 AM", "%m/%d/%Y %l:%M:%S %p")))
- 
  (as.numeric(strptime(pruned_green$lpep_pickup_datetime, "%m/%d/%Y %l:%M:%S %p")) - as.numeric(strptime("01/01/2018 12:00:00 AM", "%m/%d/%Y %l:%M:%S %p")))) /60

pruned_green <- pruned_green[trip_duration>=1.866667 & trip_duration<=46.400000,]

pruned_green %>% 
  ggplot(aes(x=trip_duration)) + 
  geom_histogram(bins=50, fill="red")+
  theme_bw()+theme(axis.title = element_text(size=12),axis.text = element_text(size=12))+
  ylab("Density")+coord_cartesian(x=c(0,50))
```


# Weather Data

Weather is closely associated with taxi demand because people are more likely to call a cab in the cold snowy winter night rather than a pleasant spring morning. Therefore, we includes the 2018 hourly weather data from https://www.ncdc.noaa.gov/cdo-web/search recorded at central park station.

```{r}
plot1 <- weather %>% 
  ggplot(aes(x=HourlyDewPointTemperature)) + 
  geom_histogram(bins=100, fill="red")+
  theme_bw()+theme(axis.title = element_text(size=12),axis.text = element_text(size=12))+
  ylab("Density")

plot2 <- weather %>% 
  ggplot(aes(x=HourlyPrecipitation)) + 
  geom_histogram(bins=100, fill="red")+
  theme_bw()+theme(axis.title = element_text(size=12),axis.text = element_text(size=12))+
  ylab("Density")

plot3 <- weather %>% 
  ggplot(aes(x=HourlyWindSpeed)) + 
  geom_histogram(bins=50, fill="red")+
  theme_bw()+theme(axis.title = element_text(size=12),axis.text = element_text(size=12))+
  ylab("Density")

grid.arrange(plot1, plot2, plot3, ncol = 3)
```
# Demand: Frequency of Pick-up

The demand is defined as the number of pick-ups in a certain time frame. In this model, we set the time frame as one hour.

```{r}
final_dt <- merge(x = pruned_green, y = weather[,-1], by.x = "pickup_hour", by.y = "HOUR")
final_dt <- merge(final_dt, final_dt[,.N, by = pickup_hour])

final_dt %>% 
  ggplot(aes(x=N)) + 
  geom_histogram(bins=1000, fill="red")+
  theme_bw()+theme(axis.title = element_text(size=12),axis.text = element_text(size=12))+
  ylab("Density") 
```


# Correlation Analysis

After carefully cleaning the data, we want to explore the correlation between each variable.

```{r}
corr_features = final_dt[,.(pickup_hour, dropoff_hour, PULocationID,RatecodeID, trip_distance,passenger_count,trip_duration,fare_amount,tip_amount,tolls_amount,total_amount,trip_type,HourlyDewPointTemperature,HourlyPrecipitation,HourlyWindSpeed,N)] 
corrplot(cor(corr_features, use='complete.obs'), type='lower')
```